---
title: "JHU COVID-19 Data Analysis Report"
author: ""
date: "2023-08-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Load required libraries
library(tidyverse)
library(knitr)
```

## Introduction

  In this report we describe four datasets on COVID-19 obtained from the Johns Hopkins University CSSE COVID-19 github site. The datasets contain information on confirmed cases of COVID-19 and COVID-19 related deaths from either the USA only, or globally. The goal of this analysis is to look at the difference in case rates and death rates from one state to another with a focus on the states of New York, Alaska, and Arizona. In addition to the comparisons across states, we wanted ask the question, can we model death rates from case rates both in the US (across states) and globally (across countries).

## Data Loading

To start, we load in the data from the four main files of time series data on COVID-19 from Johns Hopkins University. This data is obtained from the JHU CSSE COVID-19 Dataset hosted on github at the following url: https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data

```{r get_data_from_jhu}
# Get current data from the four time series files
# "time_series_covid19_confirmed_global.csv"
# "time_series_covid19_deaths_global.csv"
# "time_series_covid19_confirmed_US.csv"
# "time_series_covid19_deaths_US.csv"

url <-
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series"

filenames <- c(
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_deaths_US.csv"
)

urls <- file.path(url, filenames)

# Load population data from jhu
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

Read in the data and take a look at the structure.

```{r load_data}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2)) %>%
  rename(
    country_region = Country_Region,
    province_state = Province_State,
    population = Population
  )
```

```{r data_structure_raw, results='asis'}
kable(head(global_cases[,1:5]), caption = "global_cases data")
kable(head(global_deaths[,1:5]), caption = "global_deaths data")
kable(head(us_cases[,1:13]), caption = "us_cases data")
kable(head(us_deaths[,1:13]), caption = "us_deaths data")
```

## Data Cleaning

After reading in the four datasets, the datasets need to be tidied up to put each variable in their own column in long format. This is obvious when looking at the tables above. Additionally, there is no need for the latitude and longitude for the purpose of the planned analysis so this can be dropped from the datasets.

### Global Dataset

```{r data_cleaning_global}
# Tidy global_cases data and pivot to long format
global_cases <- global_cases %>%
  pivot_longer(cols = 
                 -c(`Province/State`,
                    `Country/Region`,
                    Lat,
                    Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))

# Tidy global_deaths data and pivot to long format
global_deaths <- global_deaths %>%
  pivot_longer(cols = 
                 -c(`Province/State`,
                    `Country/Region`,
                    Lat,
                    Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

# Join global cases and deaths datasets
global <- global_cases %>% 
  full_join(global_deaths) %>%
  rename(country_region = `Country/Region`,
         province_state = `Province/State`) %>%
  mutate(date = mdy(date))
```

The tidied "global" dataset looks much nicer now.

```{r data_structure_global, results='asis'}
kable(head(global))
```

```{r global_summary}
summary(global)
str(global)
```

It looks like there are probably a lot of rows where cases are equal to 0 so let's remove those rows.

```{r global_cleaning2}
# Keep only cases greater than zero
global <- global %>%
  filter(cases > 0)
summary(global)
str(global)

# Validate maximum number of cases
global %>% filter(cases > 103500000)
```

Here we only keep rows where cases are greater than zero and double check to make sure that the maximum values do not appear to be a typo and there is continuity in the dataset. It seems OK at this point for both cases and deaths for the "global" dataset.

### US Dataset

Next, lets clean the "US" dataset.

```{r data_cleaning_us}
# Tidy us_cases data and pivot to long format
us_cases <- us_cases %>%
  pivot_longer(cols = 
                 -c(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

# Tidy us_deaths data and pivot to long format
us_deaths <- us_deaths %>%
  pivot_longer(cols = -c(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

# Join global cases and deaths datasets
us <- us_cases %>% 
  full_join(us_deaths) %>%
  rename(county = Admin2,
         country_region = Country_Region,
         province_state = Province_State,
         combined_key = Combined_Key,
         population = Population)
```

The tidied "US" dataset looks much nicer now, but it has some additional columns which are not present in the "global" dataset. We will need to add these columns to the global dataset.

```{r data_structure_us, results='asis'}
kable(head(us), caption = "data_structure_us")
```

## Data Transformation

Here we add new columns to the "global" dataset to ensure that both the "US" and "global" datasets have the same structure.

### Adding columns to "global"

```{r add_cols_global}
# Add combined_key column to "global" dataset
global <- global %>%
  unite(
    "combined_key",
    c(province_state, country_region),
    sep = ",",
    na.rm = TRUE,
    remove = FALSE
  )

# Add population column to "global" dataset by combining with uid
global <- global %>%
  left_join(uid, by = c("province_state", "country_region")) %>%
  select(-c(UID, FIPS)) %>%
  select(province_state,
         country_region,
         date,
         cases,
         deaths,
         population,
         combined_key)
```

Now that we have the same columns in both datasets we can start the analysis.

Prior to starting our exploratory analysis we will need to create two subsets of data with summary statistics by state and summary statistics for the US totals.

### Subsetting - US by State

```{r us_by_state}
# Create us_by_state data
us_by_state <- us %>%
  group_by(province_state, country_region, date) %>%
  summarise(
    cases = sum(cases),
    deaths = sum(deaths),
    population = sum(population)
  ) %>% 
  mutate(deaths_per_mill = deaths * 1000000 / population) %>%
  select(province_state, country_region, date, cases, deaths, deaths_per_mill, population) %>%
  ungroup()
```
Let's check the structure of our subset.

```{r us_by_state_str, results = 'asis'}
# Look at data structure for us_by_state
kable(head(us_by_state), caption = "data_structure_by_state")
```
### Quality check - US by State

Now that we have our data by state, lets do a sense check on the population values to be sure everything is ok. The population of Alaska was reported to be around 731,158 in 2020 (source US Census Bureau).

```{r us_by_state_check}
# Check reported population of Alaska
filter(us_by_state, province_state == "Alaska", date == ymd("2020-01-31")) %>% select(province_state, population)
```

Here we calculate of population of 740,995 which is relatively close to the census data found online.

### Subsetting - US totals

```{r us_totals}
# Create us_totals data
us_totals <- us_by_state %>%
  group_by(country_region, date) %>%
  summarise(
    cases = sum(cases),
    deaths = sum(deaths),
    population = sum(population)
  ) %>% 
  mutate(deaths_per_mill = deaths * 1000000 / population) %>%
  select(country_region, date, cases, deaths, deaths_per_mill, population) %>%
  ungroup()
```
Let's check the structure of our subset.

```{r us_totals_str, results = 'asis'}
# Look at data structure for us_totals
kable(head(us_totals), caption = "data_structure_us_totals")
```
### Quality check - US totals

Now that we have the US totals, lets double check the population here as well. The projected total US population on January 1st, 2023 was 334,233,854 (source US Census Bureau). Our total of 332,875,137 is quite close, however, we see that the population for at the start of the pandemic is the same. As such, there may be some bias in the results knowing that the population data is static and does not change over time as it should in reality.

```{r us_totals_check}
# Check us_totals data
tail(us_totals)
```

## Exploratory Analysis

### Visualization of total data in the USA

Now that we have checked the quality of both of our US data sets, let's start by looking at our "us_totals" dataset and visualizing the data.

```{r us_totals_graph}
us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  ggtitle("COVID-19 cases and deaths in the USA") +
  labs(x = "Date", y = NULL)
```

Shown above is a graph of the total number of cases and deaths (in red and blue respectively) by date since the start of the COVID-19 pandemic in the USA. The data are plotted on a logarithmic scale to facilitate reading of the graph.

### Visualisation of data by state

Next let's look at the data across all states. For the purpose of this graph, we removed provinces/territories and only look at the 50 "official" states in the USA.

```{r all_states_graph, warning = FALSE}
us_by_state %>%
  filter(
    cases > 0,!province_state %in% c(
      "American Samoa",
      "Diamond Princess",
      "District of Columbia",
      "Grand Princess",
      "Guam",
      "Northern Mariana Islands",
      "Puerto Rico",
      "Virgin Islands"
    )
  ) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  ggtitle(str_c("COVID-19 cases and deaths across states")) +
  labs(x = "Date", y = NULL) +
  facet_wrap( ~ province_state) +
  geom_vline(xintercept = ymd("2021-06-01"), colour = "purple") +
  geom_vline(xintercept = ymd("2021-08-02"), colour = "red") +
  geom_vline(xintercept = ymd("2021-11-22"), colour = "blue")
```

From this graph we can see that the death rates and case rates vary significantly from state to state. With some states experiencing high numbers of cases much later in the pandemic than others. Several important dates in the pandemic are marked on the graph. In June 2021, the delta sub variant of COVID-19 became the dominant strain and is marked in purple. The date of August 2nd 2021 is marked in red to highlight the date where the vaccination goal of 70% of the US population vaccinated with at least one dose of the COVID-19 vaccine was met. In November 2021, the omicron sub variant of COVID-19 became was identified and is marked in blue (https://www.cdc.gov/museum/timeline/covid19.html). 

```{r new_york_graph, warning=FALSE}
state <- "New York"
us_by_state %>%
  filter(province_state == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  ggtitle(str_c("COVID-19 cases and deaths in ", state)) +
  labs(x = "Date", y = NULL)+
  geom_vline(xintercept = ymd("2021-06-01"), colour = "purple") +
  geom_vline(xintercept = ymd("2021-08-02"), colour = "red") +
  geom_vline(xintercept = ymd("2021-11-22"), colour = "blue")
```

Shown above is a graph of the total number of cases and deaths (in red and blue respectively) by date since the start of the COVID-19 pandemic in the state of New York. The data are plotted on a logarithmic scale to facilitate reading of the graph. Several important dates in the pandemic are marked on the graph. In June 2021, the delta sub variant of COVID-19 became the dominant strain and is marked in purple. The date of August 2nd 2021 is marked in red to highlight the date where the vaccination goal of 70% of the US population vaccinated with at least one dose of the COVID-19 vaccine was met. In November 2021, the omicron sub variant of COVID-19 became was identified and is marked in blue (https://www.cdc.gov/museum/timeline/covid19.html). 

At the time the preparation of this report, the latest information was from `r max(us_totals$date)` and the total number of deaths since the start of the pandemic in the USA has sadly reached `r max(us_totals$deaths)`.

## Analysis - New cases over time

After looking at the data, we are lead the the question of "Have the number of new cases leveled off?". To answer that question, we will need to go back to our data and transform it again by creating two new variables "new_cases" and "new_deaths".

### Data Transformation - new cases and new deaths

```{r add_new_cases_deaths}
# Add new_cases and new_deaths variables to us_by_state
us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

# Add new_cases and new_deaths variables to us_totals
us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```

Now that we have created the new variables, lets check what they look like for both the totals and by_state datasets.

```{r new_variables_check, results = 'asis'}
# Check us_by_state data with new variables
kable(tail(us_by_state) %>%
  select(new_cases, new_deaths, everything()), caption = "us_by_state with new variables")

# Check us_totals data with new variables
kable(tail(us_totals) %>%
  select(new_cases, new_deaths, everything()), caption = "us_totals with new variables")
```

### Visualising New Cases and New Deaths

Let's graph the new cases and new deaths over time for the US totals.

```{r us_new_cases_graph, warning=FALSE, echo=FALSE}
us_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  ggtitle("COVID-19 new cases and new deaths in the USA") +
  labs(x = "Date", y = NULL) +
  geom_vline(xintercept = ymd("2021-06-01"), colour = "purple") +
  geom_vline(xintercept = ymd("2021-08-02"), colour = "red") +
  geom_vline(xintercept = ymd("2021-11-22"), colour = "blue")

```

The graph above shows the number of new cases and new deaths (in red and blue respectively) in the USA since the start of the COVID-19 pandemic. Several important dates in the pandemic are marked on the graph. In June 2021, the delta sub variant of COVID-19 became the dominant strain and is marked in purple. The date of August 2nd 2021 is marked in red to highlight the date where the vaccination goal of 70% of the US population vaccinated with at least one dose of the COVID-19 vaccine was met. In November 2021, the omicron sub variant of COVID-19 became was identified and is marked in blue (https://www.cdc.gov/museum/timeline/covid19.html). 

```{r ny_new_cases_graph, warning=FALSE, echo=FALSE}
us_by_state %>%
  filter(province_state == "New York") %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  ggtitle("COVID-19 new cases and new deaths in the state of New York") +
  labs(x = "Date", y = NULL) +
  geom_vline(xintercept = ymd("2021-06-01"), colour = "purple") +
  geom_vline(xintercept = ymd("2021-08-02"), colour = "red") +
  geom_vline(xintercept = ymd("2021-11-22"), colour = "blue")
```
The graph above shows the number of new cases and new deaths (in red and blue respectively) in the state of New York since the start of the COVID-19 pandemic. Several important dates in the pandemic are marked on the graph. In June 2021, the delta sub variant of COVID-19 became the dominant strain and is marked in purple. The date of August 2nd 2021 is marked in red to highlight the date where the vaccination goal of 70% of the US population vaccinated with at least one dose of the COVID-19 vaccine was met. In November 2021, the omicron sub variant of COVID-19 became was identified and is marked in blue (https://www.cdc.gov/museum/timeline/covid19.html).

## Data Transformation - Case and Death rates

After analyzing the new cases and deaths, we wanted to ask, which states were the worst in terms of case rate and death rate per population. To do so, we need to go back to our data and create some new variables.

```{r add_death_case_rates}
# Add cases_per_thou and deaths_per_thou variables to us_by_state
us_state_totals <- us_by_state %>%
  group_by(province_state) %>%
  summarise(
    deaths = max(deaths),
    cases = max(cases),
    population = max(population),
    cases_per_thou = 1000 * cases / population,
    deaths_per_thou = 1000 * deaths / population
  ) %>%
  filter(cases > 0, population > 0)
```

### Analysis of highest and lowest rates

```{r check_lowest, results = 'asis'}
# Check lowest case rate states
kable(
  us_state_totals %>%
    slice_min(cases_per_thou,
              n = 10) %>%
    select(deaths_per_thou, cases_per_thou, everything()),
  caption = "Ten lowest case rate states"
)

# Check lowest death rate states
kable(
  us_state_totals %>%
    slice_min(deaths_per_thou,
              n = 10) %>%
    select(deaths_per_thou, cases_per_thou, everything()),
  caption = "Ten lowest death rate states"
)

# Check highest case rate states
kable(
  us_state_totals %>%
    slice_max(cases_per_thou,
              n = 10) %>%
    select(deaths_per_thou, cases_per_thou, everything()),
  caption = "Ten highest case rate states"
)

# Check highest death rate states
kable(
  us_state_totals %>%
    slice_max(deaths_per_thou,
              n = 10) %>%
    select(deaths_per_thou, cases_per_thou, everything()),
  caption = "Ten highest death rate states"
)
```

### Analysis of Case and Death Rates - Alaska vs. Arizona

An interesting case is that of Alaska where there is a high case rate, but a relatively low death rate. Perhaps it would be interesting to visualize the cases in Alaska vs the cases in Arizona which had a lower case rate, but a higher death rate than Alaska to better understand why that may be.

```{r ak_graph, warning=FALSE}
us_by_state %>%
  filter(province_state %in% c("Alaska", "Arizona")) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  facet_wrap(~ province_state) +
  ggtitle("COVID-19 cases and deaths in the states of Alaska and Arizona") +
  labs(x = "Date", y = NULL) +
  geom_vline(xintercept = ymd("2021-06-01"), colour = "purple") +
  geom_vline(xintercept = ymd("2021-08-02"), colour = "red") +
  geom_vline(xintercept = ymd("2021-11-22"), colour = "blue")
```

```{r ak_new_cases_graph, warning=FALSE, echo=FALSE}
us_by_state %>%
  filter(province_state %in% c("Alaska", "Arizona")) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text = element_text(angle = 90)) +
  ggtitle("COVID-19 new cases and new deaths in the states of Alaska and Arizona") +
  facet_wrap(~ province_state) +
  labs(x = "Date", y = NULL) +
  geom_vline(xintercept = ymd("2021-06-01"), colour = "purple") +
  geom_vline(xintercept = ymd("2021-08-02"), colour = "red") +
  geom_vline(xintercept = ymd("2021-11-22"), colour = "blue")
```

In the graphs above, the same important dates are marked as in previous graphs. 

There are plenty of factors that may have contributed to the higher death rate in Arizona. I would hypothesize that both the difference in age demographics and population density between Alaska and Arizona are key contributing factors, but I would need to have the age of patients for each case to test this hypothesis. Nevertheless, 13.9% of the population of Alaska is over 65 years of age vs. 18.8% in Arizona (source US Census Bureau). Considering the fact that persons over 65 years of age have a higher risk of mortality from COVID-19 infection, this may explain some of the difference in death rates seen between the two states.

  Additionally, we can see that the number of cases in Alaska remained relatively low early in the pandemic. This may be partially due to the relative isolation of Alaska and the low population density compared to Arizona. This my be a key factor contributing the the difference in death rate as the severity of disease in later variants of COVID-19 such as Delta and Omicron decreased significantly.

## Modeling COVID-19 deaths from number of cases

### US Data

Let's start of modeling our COVID-19 Data by creating a simple linear model of deaths per thousand predicted by cases per thousand using the US state totals.

```{r linear_modeling}
# Create linear model and view a summary of the linear model
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)

# Check what state has the smallest and largest cases per thousand
us_state_totals %>% slice_min(cases_per_thou)
us_state_totals %>% slice_max(cases_per_thou)

# Calculate predicted values using linear model
us_total_w_pred <- us_state_totals %>% mutate(pred = predict(mod))
```

Now that we have created a model and added the predicted values to a new dataset, we can visualise the predicted values versus the actual values.

```{r visualise_predictions}
# Plot predicted vs actual values.
ggplot(us_total_w_pred, aes(x = cases_per_thou, y = deaths_per_thou))+
  geom_point(color = "blue") +
  geom_point(aes(y = pred), color = "red")
```

We can see from the data that cases per thousand is a predictor of deaths per thousand, however, there are clearly other factors leading to differences between one state and another.

### Global data

Let's look at how this differs on a global scale. To do so, we will need to create a global totals with the deaths and cases per thousand variables.

```{r add_death_case_rates_global}
# create_global_by_country
global_by_country <- global %>%
  group_by(country_region, date) %>%
  summarise(
    cases = sum(cases),
    deaths = sum(deaths),
    population = sum(population)
  ) %>% 
  select(country_region, date, cases, deaths, population) %>%
  ungroup()

# Add cases_per_thou and deaths_per_thou variables to global
global_totals <- global_by_country %>%
  group_by(country_region) %>%
  summarise(
    deaths = max(deaths),
    cases = max(cases),
    population = max(population),
    cases_per_thou = 1000 * cases / population,
    deaths_per_thou = 1000 * deaths / population
  ) %>%
  filter(cases > 0, population > 0)

# Check structure of global_totals
head(global_totals)
```

Now that we have created out global totals dataset, we can create a linear model and look at the summary.

```{r linear_modeling_global}
# Create linear model and view a summary of the linear model
mod_global <- lm(deaths_per_thou ~ cases_per_thou, data = global_totals)
summary(mod_global)

# Check what country has the smallest and largest cases per thousand
global_totals %>% slice_min(cases_per_thou)
global_totals %>% slice_max(cases_per_thou)

# Calculate predicted values using linear model
global_total_w_pred <- global_totals %>% mutate(pred = predict(mod_global))
```

Let's visualise the predictions on the global linear model.

```{r visualise_predictions_global}
# Plot predicted vs actual values.
ggplot(global_total_w_pred, aes(x = cases_per_thou, y = deaths_per_thou))+
  geom_point(color = "blue") +
  geom_point(aes(y = pred), color = "red")
```

Here we see again that cases per thousand remains a good predictor of deaths per thousand, however, there are definitely other factors leading to the differences observed from one country to another. As mentioned previously in the Arizona vs. Alaska example, age demographics may play a role here, however, there may also be issues with how data are reported from one country to another. This leads to our final section on bias identification.

## Bias identification and Conclusions

Within the COVID-19 dataset, there may be significant differences from how deaths related to COVID-19 are recorded from one country/state to another as well as how frequently cases are recorded. This makes it challenging to compare data from one country to another and may result in bias in the datasets.

Additionally, different countries have had different access to vaccines throughout the pandemic. This may impact the relationship between cases and deaths, specifically in years following the roll out of the vaccine which was approved for emergency use in December 2020. To complicate matters further, vaccine uptake rates, mask mandates, and other COVID related restrictions varied wildly from one country/state to another.

Furthermore, following mass vaccination programs and decreasing cases globally, the quality of data on COVID-19 and the frequency of reporting has decreased overall. Data in 2023 may be relatively unreliable. With new subvariants showing a high number of mutations and possible vaccine evasion, there may be an additional wave of COVID-19 coming over the northern winter season with the need to update COVID-19 vaccines to provide protection to those who need it most.

In summary, in analyzing the US and global data sets joined by aggregating the data from JHU, on 20/08/2023 the total number of cases in the USA was  `r max(us_totals$date)` and the total number of deaths since the start of the pandemic in the USA has sadly reached `r max(us_totals$deaths)`. Globally, we have seen a total of `r sum(global_totals$cases)` cases and sadly `r sum(global_totals$deaths)` deaths since the start of the pandemic.

When comparing differences between the states, I chose the states of Alaska and Arizona for two reasons. One, I was noted that Alaska had a very low death rate despite having a high case rate overall. Arizona had a similarly high case rate, but a much higher death rate than Alaska. Two, I have family that has lived in both of those states and I thought it would be interesting to compare the two.

The differences between the two states may be due to a variety of factors, but population density, differences in age demographics, the geographical isolation and the timeline of COVID-19 in Alaska are the factors that I would hypothesize have significantly contributed to these differences.

Following the comparison of case and death rates, I then asked the question of can death rates be modeled by case rates. I built two models, one for the US only by state and one for the global dataset by country. Differences between one country and another may be due to the reasons stated above with respect to bias in the data amongst others.

As mentioned previously, COVID-19 is here to stay with a potential new wave to come this year as new variants with high levels of mutations arise. Hopefully we can learn from this pandemic to be better prepared for the next zoonotic transmission and pandemic that arises without the loss of 7 million lives globally.
